---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu

2\. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
library(readr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer 
#my file's path
swallows <- read_csv("C:/Users/Usuario/Downloads/Quiz1_ECOL596W/ECOL_596W_2024/quiz/swallows.csv")


brightness <- read_csv("C:/Users/Usuario/Downloads/Quiz1_ECOL596W/ECOL_596W_2024/quiz/brightness.csv")


# Google? y/n *no
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Answer
swallows_mrg<-merge(swallows,brightness, by="nestbox")


View(swallows_mrg)
# Google? y/n *no
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer
sum(is.na(swallows_mrg))
complete.cases(swallows_mrg)
unique(swallows_mrg$treatment)
unique(swallows_mrg$brood)
unique(swallows_mrg$n_fledged)
unique(swallows_mrg$brightness)
#to ommit NA's
swallows_mrg_clean<-na.omit(swallows_mrg)

str(swallows_mrg)
plot(swallows_mrg$brood)
plot(swallows_mrg$n_fledged)
plot(swallows_mrg$brightness)

print(swallows_mrg_clean)
# Google y/n *no
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer

#This was the first attempt
prop_fledged<-mutate(swallows_mrg_clean,(n_fledged*100)/brood)

swallows_mrg_clean$prop_fledged<-mutate(swallows_mrg_clean,(n_fledged*100)/brood)

#double checked w/ google
swallows_mrg_clean$prop_fledged<-((swallows_mrg_clean$n_fledged*100)/swallows_mrg_clean$brood)

print(swallows_mrg_clean)


# Google? y/n *yes
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 

mean(swallows_mrg_clean$treatment=="Predation")
mean(swallows_mrg_clean$treatment=="Control")

# Google? y/n *no
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer

#first I cleaned the NA's
swallows_mrg_clean2<-na.omit(swallows_mrg_clean)
unique(swallows_mrg_clean2$prop_fledged)
unique(swallows_mrg_clean2$treatment) #double checked
swallows_mrg_clean2$treatment <- as.factor(swallows_mrg_clean2$treatment)
swallows_mrg_clean2$prop_fledged <- as.numeric(swallows_mrg_clean2$prop_fledged)


?boxplot        

#first look
boxplot(swallows_mrg_clean2$treatment,swallows_mrg_clean2$prop_fledged)

#my attempt
plot(swallows_mrg_clean2$treatment,swallows_mrg_clean2$prop_fledged, 
     main = "Fledging success",
     xlab = "Treatment",
     ylab = "Fledging %",
     col="orange")

#I I had a mistake and I corrected it with google
boxplot(swallows_mrg_clean2$prop_fledged ~ swallows_mrg_clean2$treatment,
        main = "Fledging success",
        xlab = "Treatment",
        ylab = "Fledging %",
        col = "orange")

# Google? y/n *yes
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer

#first look
plot(swallows_mrg_clean2$brightness,swallows_mrg_clean2$prop_fledged)

#my attempt
library(ggplot2)
swallows_mrg_clean2 |> ggplot(aes(x=prop_fledged, y=brightness))+
  geom_smooth(
    color="purple", fill="lightpink") +
  labs(title = "Influence of female brightness on fledging success",
       x= "Fledging %",
       y="Brightness") 
  

# Google? y/n **here I used google to ask how to change the "fill" color
```
