---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(ggplot2)
library(dslabs)
library(dplyr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer 
<<<<<<< HEAD
 swallows<-read.csv("swallows.csv", header = TRUE)
 brightness<-read.csv("brightness.csv", header = TRUE)

# Google? y/n
 n
=======

# Google? N

>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.undefined

```{r}
#Answer
<<<<<<< HEAD
comb_data <- swallows %>%
  left_join(brightness, by = "nestbox")
# Google? y/n
n
=======


# Google? Y

>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer
<<<<<<< HEAD
str(comb_data)
summary(comb_data)
colSums(is.na(comb_data))
comb_new <- comb_data %>%
  filter(!is.na(brightness))
# Google y/n
#y for NA values
=======
# Google N

>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer
<<<<<<< HEAD

comb_new2 <- comb_new %>%
  mutate(prop_fledged = ifelse(brood>0, n_fledged/brood, NA))

# Google? y/n
n
=======
# Google? yes. extensively. this took me 30 minutes since it wouldn't let me use the tidyverse mutate
>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 
<<<<<<< HEAD
comb_new2 %>%
  group_by(treatment) %>%
  summarize(mean_fledged = mean(prop_fledged, na.rm = TRUE))


# Google? y/n
y
=======

# Google?
>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

Create a plot showing fledging success differences between treatments

```{r}
<<<<<<< HEAD
#Answer
ggplot(comb_new2, aes(x = treatment, y = prop_fledged)) +
  geom_boxplot() +
  labs(x = "treatment", y = "proportion of fledged birds", 
       title = "fledging success differences between treatments")

# Google? y/n
n
=======
# Answer
# Google? N

>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer
<<<<<<< HEAD
# Convert a column to numeric
comb_new2$brightness<- as.numeric(comb_new2$brightness)

correlation <- cor.test(x = brightness, y = prop_fledged, method = "pearson")
print(correlation)

ggplot(data.frame(comb_new2), aes(x = brightness, y = prop_fledged)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

# Google? y/n
n
=======
#Google?
>>>>>>> e9ed7c218b5f16ad1e5e2dd79299ac6bf25301d5
```
