---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
<<<<<<< HEAD
# Answer 
swallows <- read.csv("swallows.csv")
brightness <- read.csv("brightness.csv")

# Google? N
=======
# Answer
#I used the "open data" function on R
 


>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.undefined

```{r}
#Answer
<<<<<<< HEAD
swallows_updated <- merge(swallows, brightness, by = "nestbox")
#I would check for any NAs (if for whatever reason the dimensions didn't line up, due to the original df or something I did). Additionally, I can manually check a sample of rows, comparing between _updated and the two original to ensure everything worked as expected.
# Google? Y
#I initially tried in tidyverse with mutate (hence Googling), but got nowhere fast, so I dropped back to base R w/ no Googling
=======

#I used total <- merge(brightness,swallows,by="nestbox"). I checked by saying View(total)
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
<<<<<<< HEAD
# Answer
str(swallows_updated)
summary(swallows_updated$brood)
summary(swallows$brood)
#Summary looks good here...
summary(swallows_updated$n_fledged)
summary(swallows$n_fledged)
summary(swallows_updated$brightness)
summary(brightness$brightness)
#Of course, I have no idea what the units are for brightness, so I'm going to assume these numbers are sane. They align, at least.
# Google N
=======
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer
swallows_updated <- mutate(swallows_updated, prop_fledged=n_fledged/brood)
#Hmmm... getting some NaN but I'm going to choose to ignore, but be aware of it, for now.

<<<<<<< HEAD
# Google? N
=======
#total$prop_fledged <- with(total, brood / n_fledged). I then tidied it up by removing the Na and replacing INF with )

# Google? yes. extensively. this took me 30 minutes since it wouldn't let me use the tidyverse mutate
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 
#Never mind, I'm going to have to be aware of NaN now and exclude them. Sigh...
swallows_updated %>% group_by(treatment) %>% summarize(mean = mean(prop_fledged, na.rm=TRUE))

<<<<<<< HEAD
#Control = 0.70555555...
#Predation = 0.335; those make sense
# Google? N
=======
#the mean proportion of nestlings fledged was 1.484

#I used summary(total) to get the different stats for each column, had to rename as "View(na.omit(total))" 

# Google? not this time!
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer
library(ggplot2)
ggplot(swallows_updated, aes(x=treatment, y=prop_fledged)) +geom_boxplot()

<<<<<<< HEAD
# Google? N
=======
#ggplot(na.omit(total))
> ggplot(data = na.omit(total), mapping = aes(treatment, n_fledged))
> ggplot() +
+     geom_point(mapping = aes(x = treatment, y = n_fledged), data = na.omit(total)) +
+     geom_point(
+         mapping = aes(x = treatment, y = n_fledged), data = na.omit(total),
+         colour = 'red', size = 3
+     )
> 
# Google? no!
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer
#That's actally a really interesting question for someone who has no idea how birds work...
ggplot(swallows_updated, aes(x=brightness,y=prop_fledged)) +geom_smooth(method="lm") +geom_point()
#Doesn't appear to be affected!

<<<<<<< HEAD
# Google? Y, for the smoothening function, specifically about the method argument.
=======
#Female brightness seems to matter on both ends: darker females and lighter females along the spectrum will have more chicks that fledge, but females in the middle will see a decrease in the number of chicks successfully fledged 

#first I made a scatterplot using this: ggplot() +
+     geom_point(mapping = aes(x = brightness, y = n_fledged), data = na.omit(total)) +
+     geom_point(
+         mapping = aes(x = brightness, y = n_fledged), data = na.omit(total),
+         colour = 'red', size = 3

#Then I added a trend line to see if it mattered by doing this: ggplot(na.omit(total), aes(x=brightness, y=n_fledged)) + geom_point() + geom_smooth()


# Google? yes, so much!
>>>>>>> 26677ef8a1d7fa4dde9984fd3d518fc73b6a0600
```
