---
title: "cookie PCAs"
author: "Sabrina McNew"
date: "2023-11-21"
output: html_document
---
# Coookie stuff

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages and data
```{r, echo=FALSE}
devtools::install_github("vqv/ggbiplot")
library(ggbiplot)
library(dplyr)
cookies <- read.csv("datasets/cookies.csv")
```

The cookies dataset has 214 recipes. I cleaned it up so that just the main/common
ingredients are represented. Obviously there are other sources of variation that 
affect cookie product (preparation method, cook time, elevation, etc) but let's 
just look at variation among ingredients. 

```{r}
head(cookies)
```

Can we use a PCA to identify emergent properties of these different recipes? 
Create the PCA object. Remember that you need to only include numeric columns in the PCA
```{r}
pca1 <- cookies %>% select(-c(X, Recipe_Index)) %>% 
                             prcomp(x = ., center = TRUE, scale = TRUE)
```

What's in the PCA object? lots of stuff
```{r}
str(pca1)

```
Let's look at the summary. There are 13 ingredients so there are 13 PCs. 
PC1 explains 32% of overall variation
```{r}
summary(pca1)
```

What about the ingredients? How do they "load" onto the PCs?
```{r}
pca1$rotation
```
### Let's make a plot
This is sort of hard to parse, because the vectors are going all over the place. 
Can we label the recipes we made? 
```{r}
ggbiplot(pca1,
         obs.scale = 1,
         var.scale = 1)
```
```{r}
# I'm going to add PC1 and PC2 back on to my original dataframe 
pca_values <- pca1$x %>% as.data.frame# this is a dataframe containing the PC values
str(pca_values)
cookies <- cookies %>% mutate(PC1 = pca_values$PC1,
                              PC2 = pca_values$PC2)

# I am going to plot PC1 vs. PC2, and then label just the recipes we focused on
cookies %>% ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_text(data = filter(cookies, Recipe_Index %in% c("Bon_appetit", "joy", "betty", "H_A", "yum")),
            aes(PC1, PC2, label = Recipe_Index, color = "red", size = 14))

```

\
