---
title: "Dslabs Chapter 8 answers"
author: "Sabrina McNew"
date: "2023-09-14"
output: html_document
---
# R. Irizarry Introduction to Data Science
## Chapter 8 Exercises 
### 8.15 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Step 1, load data and relevant packages
```{r, message = FALSE}
library(dplyr)
library(ggplot2)
library(dslabs)
data(heights)
data(murders)
```

```{r}
p <- murders |> ggplot()
class(p)
```
Class of this plot is "gg" and "ggplot." Sabrina's note: I did not
know this. 

```{r}
p
```
2. Remember that to print an object you can use the command print or simply type the object. Print the object p defined in exercise one and describe what you see.
  
"p" is a blank plot 

```{r}
p <- heights |> ggplot()
class(p) #same thing
head(murders)
```

"total" is the column for murders and "population" is the col for pop size

6. To create the scatterplot we add a layer with geom_point. The aesthetic mappings require us to define the x-axis and y-axis variables, respectively. So the code looks like this:

```{r}
murders |> ggplot(aes(x = total, y = population)) +
  geom_point()
```
7. Note that if we don’t use argument names, we can obtain the same plot by making sure we enter the variable names in the right order.

Sabrina note: for all R functions, you don't have to say arg = 
if you put the arguments in the right order. 
so:

```{r, eval = F}
murders |> ggplot(aes(population, total)) +
  geom_point() 
```
will work just as well as 
```{r, eval = F}
murders |> ggplot(aes(x = population, y = total)) +
  geom_point()
```
**however** it is good practice to always include your 
argument names (i.e. "x = ", "y = ", "data = ") for readability
and to make sure you and R are on the same page. 

8. If instead of points we want to add text, we can use the geom_text() or geom_label() geometries. The following code
murders |> ggplot(aes(population, total)) + geom_label() will give us an error message. Why?

I think the answer is: 
"We need to map a character to each point through the label argument in aes."

9. Rewrite the code above to use abbreviation as the label through aes
```{r}
murders |> ggplot(aes(x = population, y = total, label = abb)) + geom_label()
```
# Fun fact, you can also define aesthetics within geoms. 
```{r}
murders |> ggplot(aes(x = population, y = total)) + 
  geom_label(aes(label = abb))
```
10. Change the color of the labels to blue. How will we do this?
Answer: d works.
```{r}
murders |> ggplot(aes(x = population, y = total, label = abb)) + 
  geom_label(color = "blue")
```

11. see above

12. Now suppose we want to use color to represent the different regions. In this case which of the following is most appropriate:

Answer: -> Because each label needs a different color we map the colors through the color argument of aes .

```{r}
murders |> ggplot(aes(x = population, 
                      y = total, 
                      label = abb,
                      color = region)) + 
  geom_label()
```
14. Now we are going to change the x-axis to a log scale to account for the fact the distribution of population is skewed. Let’s start by defining an object p holding the plot we have made up to now
```{r}
p <- murders |> 
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label() 
p + scale_x_log10()
```

15. Repeat the previous exercise but now change both axes to be in the log scale.
```{r}
p <- murders |> 
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label() 
p + 
  scale_x_log10() +
  scale_y_log10()
```
16. Now edit the code above to add the title “Gun murder data” to the plot. Hint: use the ggtitle function.

```{r}
p <- murders |> 
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label() 
p + 
  scale_x_log10() +
  scale_y_log10() +
  ggtitle("Gun murder data")
```
### Sabrina's note: 
This still looks kind of ugly. Here's an improvement
```{r, warning = FALSE }
library(ggthemr)
ggthemr(palette = "solarized", layout = "clean", text_size = 14)

options(scipen = 999) #supress scientific notation
murders |> 
  ggplot(aes(population/1000000, total, label = abb, color = region)) +
  geom_text() +
  scale_x_log10() +
  scale_y_log10() +
  ggtitle("Gun murder data") +
  labs(y = bquote("Murders per year (" ~ log[10] ~ "scale)"),
       x = bquote("Population (millions," ~ log[10] ~ "scale)"),
       color = "Region")

```

