---
title: "Dslabs Answer Key Ch. 4"
author: "Sabrina McNew"
date: "2023-09-07"
output: html_document
---

### Load our libraries and set plot theme
```{r, warning = FALSE, message = FALSE}
library(dslabs)
library(ggthemr)
library(tidyr)
library(dplyr)

ggthemr(palette = "flat", layout = "clean", text_size = 22)
```

## Dslabs: Introduction to Data Science Book First edition
### Rafael Irizarry

## Chapter 4

### 4.2 Exercises
1. Examine the built-in dataset co2. Which of the following is true:
2. Examine the built-in dataset ChickWeight. Which of the following is true:
3. Examine the built-in dataset BOD. Which of the following is true:
4. Which of the following built-in datasets is tidy (you can pick more than one):

```{r}
head(co2) # vector of numbers
head(ChickWeight)
head(BOD)

```
Sabrina's notes:  
1. I guess this is *b*; it's just a vector so I wouldn't say it's "tidy." 
I don't think you really need a character column to be tidy, but if it were 
a dataframe it might be tidier. Not sure about this q tbh.  

2. *b* This seems pretty tidy. It's ok that individual chicks have more than one
measurement.  

3. *c* looks fine  

4. Ones that are not tidy:  
BJSales, is just a vector, not sure what other data relate to it. 
EuStockMarkets: some sort of strange class ("time series") seems more like 
a matrix. I'm going with untidy, though probably someone wanted it in this format
UCBAdmissions: terrible format; List of tables, not good. 

### 4.4 Exercises
1. Use the function mutate to add a murders column named rate with the per 100,000 murder rate as in the example code above. Make sure you redefine murders as done in the example code above ( murders <- [your code]) so we can keep using this variable.

```{r}
murders <- mutate(murders, rate = total / (population/100000))
```
2. If rank(x) gives you the ranks of x from lowest to highest, rank(-x) gives you the ranks from highest to lowest. Use the function mutate to add a column rank containing the rank, from highest to lowest murder rate. Make sure you redefine murders so we can keep using this variable.
```{r}
murders <- mutate(murders, rank = rank(-rate))

```

3. Use select to show the state names and abbreviations in murders. Do not redefine murders, just show the results.

```{r}
murders |> select(state, abb) %>% head
```

4. Use filter to show the top 5 states with the highest murder rates. After we add murder rate and rank, do not change the murders dataset, just show the result. Remember that you can filter based on the rank column.
```{r}
murders |> filter(rank %in% c(1:5))
```

5. Create a new data frame called no_south that removes states from the South region. How many states are in this category? You can use the function nrow for this.
Sabrina Note: to save time/coding lines I'm opting just to pipe the result to nrow()
to get the number rather than creating a new data farme and assigning it to the namespace

```{r}
murders |> filter(region != "South") |> nrow()
```
6. Create a new data frame called murders_nw with only the states from the Northeast and the West. How many states are in this category?

```{r}
murders |> filter(region %in% c("Northwest", "West")) |> nrow()
```

7. Create a table called my_states that contains rows for states satisfying both the conditions: it is in the Northeast or West and the murder rate is less than 1. Use select to show only the state name, the rate, and the rank.

```{r}
murders |> filter(rate < 1 & region %in% c("Northwest", "West")) |>
  select(state, rate, rank)
```
### 4.6 Exercises

1.Repeat the previous exercise, but now instead of creating a new object, show the result and only include the state, rate, and rank columns. Use a pipe |> to do this in just one line. (Answer: see above, already piped for convenience)

2. Reset murders to the original table by using data(murders). Use a pipe to create a new data frame called my_states that considers only states in the Northeast or West which have a murder rate lower than 1, and contains only the state, rate and rank columns. The pipe should also have four components separated by three |>. The code should look something like this:
```{r}
data(murders)
my_states <- murders |> 
  mutate(rate = total / (population/100000),
         rank = rank(-rate)) |> 
  filter(region %in% c("Northeast", "West")) |> 
  filter(rate < 1) |> 
  select(state, rate, rank)
# Sabrina note, I personally like to separate my filters but YMMV
```

### 4.10 Exercises
```{r, message = FALSE}
#install.packages("NHANES")
library(NHANES)
data(NHANES)
```

1. We will provide some basic facts about blood pressure. First let’s select a group to set the standard. We will use 20-to-29-year-old females. AgeDecade is a categorical variable with these ages. Note that the category is coded like ” 20-29”, with a space in front! What is the average and standard deviation of systolic blood pressure as saved in the BPSysAve variable? Save it to a variable called ref.

Sabrina's notes: I interpreted ref should be the systolic blood pressure vector

```{r}
head(NHANES) #let's have a look 
ref <- NHANES |> 
      filter(AgeDecade == " 20-29" & Gender == "female")  |> 
      pull(BPSysAve)
mean(ref, na.rm = T)
sd(ref, na.rm = T)
```
2. Using a pipe, assign the average to a numeric variable ref_avg. Hint: Use the code similar to above and then pull.
```{r}
ref_avg <- NHANES |> 
      filter(AgeDecade == " 20-29" & Gender == "female")  |> 
      pull(BPSysAve) |> 
      mean(na.rm = T)
ref_avg
```

3. Now report the min and max values for the same group.
```{r}
NHANES |> 
      filter(AgeDecade == " 20-29" & Gender == "female")  |> 
      pull(BPSysAve) |> range(na.rm = T)
```
4. Compute the average and standard deviation for females, but for each age group separately rather than a selected decade as in question 1. Note that the age groups are defined by AgeDecade. Hint: rather than filtering by age and gender, filter by Gender and then use group_by.
```{r}
NHANES |> 
  filter(Gender == "female") |> 
  group_by(AgeDecade) |> 
  summarize(mean_bps = mean(BPSysAve, na.rm = T),
            sd_bps = sd(BPSysAve, na.rm = T))
```
5. Repeat exercise 4 for males.

```{r}
NHANES |> 
  filter(Gender == "male") |> 
  group_by(AgeDecade) |> 
  summarize(mean_bps = mean(BPSysAve, na.rm = T),
            sd_bps = sd(BPSysAve, na.rm = T))
```

6. We can actually combine both summaries for exercises 4 and 5 into one line of code. This is because group_by permits us to group by more than one variable. Obtain one big summary table using group_by(AgeDecade, Gender).
```{r}
NHANES |> 
  group_by(AgeDecade, Gender) |> 
  summarize(mean_bps = mean(BPSysAve, na.rm = T),
            sd_bps = sd(BPSysAve, na.rm = T))
```
