---
title: 'Week 8 Exercises: GLM, LMM, and GLMM'
author: "Sabrina McNew"
date: "2023-10-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message = F}
# Packages used (install as needed)
library(ggplot2)
library(ggthemr)
library(lmerTest) # for extracting p vals from mixed models
library(lme4) # mixed models
library(dplyr)
```
```{r}
# datasets used
nestlings <- read.csv("../datasets/nestlings.csv")
swallows <- read.csv("../datasets/swallows.csv")

# optional set theme
ggthemr(palette = "flat dark", layout = "clean", text_size = 18)
```

###Question 1:  
In the swallows dataset, each row is a breeding female tree swallow
and the columns tell you her brightness, which treatment she was in, her brood size,
and her breeding success. There are two ways to consider breeding success:
either as a binary "did she fledge any young or not?" encoded in the "success"
column, or by considering the number of nestlings that lived and died per female
("nestlings_lived" and "nestlings_died" columns).

1a. Create a plot of the proportion of successful breeders in the predation
group and in the control group.
```{r}
swallows %>% 
  ggplot(aes(x = treatment, fill = as.factor(success))) + 
  geom_bar(position = "fill")

# another option if we just want to visualize success:
swallows %>% 
group_by(treatment) %>% 
summarize(proportion = mean(success)) %>% 
  ggplot(aes(x= treatment, y = proportion)) + 
  geom_col()

```


1b. Create a plot of the overall percentage of nestlings that fledged in
the predator and control groups.

```{r}
head(swallows)

#I'm old so I'm going to use aggregate in base r. summarize in dplyr will also work
bird_sums <- aggregate(cbind(nestlings_lived, brood_size) ~ treatment, data = swallows, FUN = sum)

bird_sums %>% 
  mutate(percent_fledge = nestlings_lived/brood_size) %>% 
  ggplot(aes(x = treatment, y = percent_fledge)) +
  geom_col()
```

1c. Use a generalized linear model (aka logistic regression) to test whether
breeding success differed between predator and control groups using "success"
as a response variable.
```{r}
glm(success ~ treatment, data = swallows, family = "binomial") %>% summary()
```
Sabrina's note: See that this will run, and it will give you the "right" 
answer. It's pretty easy to draw a line through things and it is interpreting
that there's more "1s" in the control group than in the treatment group. R 
is not going to tell you that you're using the "wrong" model, that's up to you.
But, it is reassuring to know that different models will come to the same 
conclusion.

1d. Use a generalized linear model (aka logistic regression) to test whether
the proportion of nestlings that fledged from each nest differed between
treatments. How do you interpret the output?

```{r}
glm(cbind(nestlings_lived, nestlings_died) ~ treatment, data  = swallows, 
    family = "binomial") %>% summary()
```


1e. Use a generalized linear model to test whether the total number of
offspring that females succesfully produced differed between treatment and control groups.
What glm family should you use?

```{r}
glm(nestlings_lived ~ treatment, data = swallows, family = "quasipoisson") %>% summary()

mean(swallows$nestlings_lived)
var(swallows$nestlings_lived)
```
Sabrina's notes: Number of fledglings per nest is something that might feel pretty poisson-y. Again, beause I know it's never poisson, I'm going to use a relative, in this case "quasipoisson." In this case, when we look at the mean vs. var of nestlings_lived, it's not *that* over-dispersed. But, there's no real harm here in using an alternative.

One other thing to maybe note: in cases like this we may run into what's called "zero inflated data." That's when we have more zeros than we'd expect, 
for example, because some biological process is generating complete nest failures (like predation). Then, a separate process generates variation fledging success among those nests pass the baseline of "not failing." There are some other packages that are useful for modeling zero inflated data and other flavors of not-poisson. But save that for another day... 

1f: Considering your approaches in 1c - 1e: what is the "best" way to answer
the question: "Did predation affect breeding success of swallows?" Why?

?????? Up to you! They arguably measure different things:  
1) Did a female bird have any reproductive success or not (success ~ treatment)  
2) What proportion of nestlings fledged in each treatment? (cbind(lived,died) ~ treatment)   
3) Did total fitness depend on treatment? (lived ~ treatment)

### Question 2:   
You also have data from each nestling in this experiment in the
"nestlings" dataframe, which includes measurements at 12 days of age of their
mass and their wing length. You want to determine whether being in a "predation" or "control" treatment had an effect on individual nestling mass. So you set out to model the effect of treatment on nestling mass using lm(). "HOLD UP" your advisor says, that's pseudoreplication, because each nestling is not independent. "What you need to do is take the average mass for each nest and then model average mass based on treatment"
"But," you say, "that will sacrifice a lot of power! Maybe there's another way..."

2a. Make a plot of mean mass for the predator and control treatments
```{r}
nestlings |> ggplot(aes(x = treatment, y = mass)) + geom_boxplot()
```

2b. Make a lm that tests the effect of treatment on nestling size. How do you
interpret the coefficients and p vals?
```{r}
lm(mass ~ treatment, data = nestlings) |> summary()
```
The bare-bones lm is pretty confident that nestlings in the Predation 
treatment were smaller than the Controls (P = 0.0085). Specifically, nestlings
in the control group weighed ~ 18g (on day 12) while nestlings in the predation
group weighed ~ 17g (18.5 - 1.09) at the same age. 

2c. Make a lmm that tests the effects of treatment on nestling size, properly
accounting for the interdependence of nestlings within the same nest. (hint: lmer()).
Compare and contrast the results of your lm and lmm. How do the coefficients and
p values differ?

```{r}
lmer(mass ~ treatment + (1|nest), data = nestlings) |> summary()
```
### Semi-important note I didn't say in class: 
It is very hard to calculate p values for LMMs because it's hard to decide what 
the dfs are. Our rule of thumb is that dfs = N - number of estimated parameters. 
But what is N for a mixed model? It is not the number of nestlings (because of 
pseudoreplication), and it's not exactly number of nests either (because we do 
have more information in our data set). It's somewhere in the middle. 

In order to extract P values, you have to make a decision about these dfs, and lme4 
won't do it on its own, because it does require some assumptions and is somewhat 
subjective. What lmerTest does is chooses a way to get a P value (the Satterthwaite method)
and it runs it. But the writers of lmer4 didn't want you sail by that point and they wanted you
to realize it's not trivial. In practice, using lmerTest is fairly common, 
and if you were using another stats program (SPSS etc.) in the fine print you'd read
that they're also using an "approximation" of some sort to get to the p vals for LMM. 

###Question 3:  
Test for effects of treatment on individual nestling fledging success
3a: Make a plot showing the differences in percent fledging success for nestlings
in the control vs. predator treatment

```{r}
nestlings %>% ggplot(aes(x = treatment, fill = nestling_fate)) + geom_bar(position = "fill")
```


3b: Use a normal lm to model the effects of treatment on fledging success (yes this isn't the best approach). You'll first want to create a new vector that codes nestling fate as a numeric
value. How do you interpret these coefficients? Do they make sense?
```{r}
# note this way of adding a column isn't super readable but it works just fine. 
# Another way to do it could be mutate(fledge_numeric = case_when(nestling_fate == "Fledged" ~ 1))
nestlings |> 
  mutate (fledge_numeric = as.numeric(as.factor(nestling_fate))) %>% 
  lm(fledge_numeric ~ treatment, data = .) %>%
  summary()

# another note: because I'm moving quickly and not interested in saving 
# a bunch of new objects to my environment I'm just going to pipe these things together. 
# writing it this way will not preserve the new column, but that's ok with me. 
```
## Super annoying and concerning error:  
I've been trying to force myself to use the native pipe |> instead of the tidyverse
pipe %>% recently because it seems like the new cool thing. However, I just found 
an example of where the |> pipe throws an unexpected error: 
```{r, eval=FALSE}
# New pipe
nestlings |> 
  mutate (fledge_numeric = as.numeric(as.factor(nestling_fate))) |>
  lm(fledge_numeric ~ treatment, data = _) # destination indicated by _

# Old pipe
nestlings |> 
  mutate (fledge_numeric = as.numeric(as.factor(nestling_fate))) %>%
  lm(fledge_numeric ~ treatment, data = .) # destination indicated by .
```
Thanks to google I realized that if you want to set a destination 
for your piped object in old piping you use "." and in new piping you use "_". So that's super annoying to me but now we know. It also appears that %>% can do a few things 
that |> cannot. Read more here: https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/
It is generally good practice to use Base R functions when you can, because 
dependencies make things more complicated. However, ymmv with pipes- do what makes sense 
for your workflow. 

3c: Use a generalized linear model aka logistic regression (glm(family = "binomial)) to model the effect of treatment on nestling fledging success. How do you interpret the coefficients?

Tip: R will not allow a character vector as a response in a logistic regression.
however, it will take a 0 or 1 number, and it will take a vector encoded as a two-level factor

### going back to my old pipe 
```{r}
glm(as.factor(nestling_fate) ~ treatment, data = nestlings, family = "binomial") %>%
  summary
```

The p value here (0.034) tells us that the probability than an 
individual nestling fledges is smaller in the predation group. 

3d: Use a generalized linear mixed effects model to model the effect of treatment on nestling fledging success, accounting for the interdependence of nestlings. Interpret these results. Compare the results of the lm, glm, and glmm.

```{r}
glmer(as.factor(nestling_fate) ~ treatment + (1|nest), data = nestlings, family = "binomial") %>%
  summary()
```
Once we make the "right" model R is pretty confident that the mass
of nestlings is smaller in the predation treatment. 


### Question 4:   
The swallow nestlings were studied at 5 different sites, which varied in temperature, proximity to food sources, and overall quality. Site therefore may be an important random effect to consider in our analysis.

4a: Inspect nestlings$site. How many sites are there? How many nests were in each site? Make a plot of average nestling mass at each site. How does it vary?

Sabrina's note: sorry, should have said "how many nestlings at each site." 
```{r}
table(nestlings$site) # five sites, most nestlings at unit 4
nestlings %>% 
  ggplot(aes(x = site, y = mass)) + geom_boxplot() 
```
Interesting, nestling mass seems to vary quite a bit among sites 
4b: Create a plot of the relationship between wing length and mass, coloring
points by site. Add a simple regression line for the points of each site on top
of your graph.
```{r, warning = F}
nestlings %>% 
  ggplot(aes(x = wing, y = mass, color = site)) +
  geom_point() +
  geom_smooth(method = "lm")
```


4c: Model the relationship between wing length and mass using a simple lm
```{r}
lm(mass ~ wing, data = nestlings) %>% summary() # super significant
```

4d: Model the relationship between wing length and mass using the random effect
of site.

```{r}
lmer(mass ~ wing + (1|site), data = nestlings) %>% summary

```
Site does explain some variation in mass, but wing still explains a lot. 

4e: What other random effects should you consider? Experiment with adding the random
effect of nest. How should you add it? For a nice explanation of nested vs.
crossed random effects read here: https://stats.stackexchange.com/questions/228800/crossed-vs-nested-random-effects-how-do-they-differ-and-how-are-they-specified
```{r}
mod1 <- lmer(mass ~ wing+ (1|year) + (1|site/nest), data = nestlings)
summary(mod1)
plot(mod1) #Plot of residuals looks basically fine

# See if a random slope will work. I won't do it for year, bc only two years. 
# By putting "wing" in front of our random effects, we are asking R to say not only
# could the intercept (i.e. average mass) differ among sites and/or nests 
# But the relationship betwene wing and mass (i.e. the slope) may also 
# vary among sites and nests. 
mod2 <- lmer(mass ~ wing+ (1|year) + (wing|site/nest), data = nestlings)
summary(mod2)
```

Note, this does give a convergence warning (boundary (singular) fit). This error
is common as your random effects become more complex. It doesn't mean that your 
model is wrong per se, but it does suggest that you don't have quite enough
data to parameterize it. If you get a convergence error, common guidance is to 
simplify your random effects structure. 

### Final notes: 
Random effects models can get tricky, especially when you start specifying 
slopes and intercepts, and especially if you get into the weeds of p values. 

**If this is your first rodeo with GLMs LMMS: ** Try to familiarize yourself 
with how the models run and how they're interpreted. Try to understand when you'd 
use a glm vs. a lmm, and how the syntax works. If you are having trouble with 
any of that talk to Sabrina. 

**If you feel pretty confident with GLMs LMMs:** Expand your knowledge 
by reading up on some of the technical details and in-depth considerations 

Ben Bolker is the LMM god, so read his stuff and listen to whatever he says 
on Stack Overflow. 
https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#what-are-the-p-values-listed-by-summaryglmerfit-etc.-are-they-reliable

Other stats people's explanation of random effects: 
https://bookdown.org/steve_midway/DAR/random-effects.html
https://m-clark.github.io/mixed-models-with-R/random_slopes.html


