---
title: "Quiz 1 Study guide"
format: html
editor: visual
---

## Format of the quiz

### Part 1: Reading Comprehension 
You will get a paper document with R language on it, you will be asked to 
interpret and explain in plain English what the code means. No use of Google or
R help pages please. 


### Study guide: 

Base R Operators 

```{r, eval = F}
# Know what the following mean
<-
()
[]
:
!
-
$
```

* Understand how to select and add columns to a data frame 
* How to select particular rows and columns in a data frame 
* The common functions that we use to do sanity checks on data 
(e.g., table(), str(), summary())
* Common classes of variables 
* The order in which functions are evaluated in R

Dplyr operators 
```{r, eval = F}
# Know the following in dplyr/tidyverse
# Pipes and how they work 
%>% and |> 

```

* Understand how the major dplyr wrangling functions work (mutate, filter,
select, pull, summarize)
* Understand how expressions are evaluated using pipes 


### Part 2: Coding in action
The quiz will be posted in ECOL_596W_2024/Assignments. It will be a quarto document
that you should download and modify on your own computer. Submit your answers 
via a pull request or email. 

### Study guide: 

Be able to load, wrangle, summarize, and plot data. 

* Everything in the quiz you've done multiple times in problem sets
* You can use base R or tidyverse syntax 
* Definitely (always) use the help files
* Google and ChatGPT are allowed, but *please* disclose if you use them 

