---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
library(ggplot2)
library(ggthemr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer 
 #brightness <- read_csv("C:/Users/allys/OneDrive/Documents/Ally Learns R/myrepo/ECOL_596W_2024/ECOL_596W_2024/quiz/brightness.csv")
#swallows <- read_csv("C:/Users/allys/OneDrive/Documents/Ally Learns R/myrepo/ECOL_596W_2024/ECOL_596W_2024/quiz/swallows.csv")
brightness <- read.csv("../quiz/brightness.csv")
swallows <- read.csv("../quiz/swallows.csv")
 # Google? n
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Answer
data <- left_join(swallows, brightness, by = "nestbox")
# Google? y
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer
str(data) #look at data, object types look pretty good to me

missing_values <- colSums(is.na(data))
print(missing_values) #check for NAs 


c_data <- data %>% filter(!is.na(brightness)) %>% filter(brood > 0) # remove those NAs & nests with no babies



# Google n
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer
c_data$prop_fledged <- c_data$n_fledged/c_data$brood 


# Google? n
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 
grouped_data<- c_data %>%
  group_by(treatment) %>%
    summarize(mean_prop_fledged = mean(prop_fledged, na.rm = TRUE))

# Google? n
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer
grouped_data %>%
  ggplot(aes(x = treatment, y = mean_prop_fledged)) +
    geom_bar(stat = "identity") +
    labs(title = "Mean Proportion of Fledged Nestlings by Treatment Group",
         x = "Treatment Group",
         y = "Mean Proportion Fledged") +
    theme_minimal()

# Google? n
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer
c_data %>%
  ggplot(aes(x = brightness, y = prop_fledged)) +
    geom_point(alpha = 0.7) + 
    geom_smooth(method = "lm", se = FALSE, color = "blue") + 
    labs(title = "Relationship Between Female Brightness and Fledgling Success",
         x = "Female Brightness",
         y = "Proportion Fledged") +
    theme_minimal()

#It appears to me that there is no relationship between female brightness and fledging sucess. 
# Google? y
```
