---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

\## Part 2: Coding in action

\### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data
library(dslabs)
library(dplyr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer
swallows <- read.csv('../quiz/swallows.csv')
brightness <- read.csv('../quiz/brightness.csv')

# Google? y/n [n]
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Answer

bright_swa <- merge(swallows, brightness, by='nestbox')

# Google? y/n [y]
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer

bright_swa <- na.omit(bright_swa)

# Google y/n [n]
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer

bright_swa <- bright_swa %>%
  mutate(prop_fledged = n_fledged/brood) %>%
  mutate_all(~replace(., is.nan(.), 0))


# Google? y/n [y for dropping NaNs]
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer

bright_swa %>%
  group_by(treatment) %>%
  summarize(average = mean(prop_fledged))

# Google? y/n [n]
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer

library(ggplot2)
fledg_plot <- bright_swa %>%
  ggplot(aes(treatment, prop_fledged)) +
  geom_violin()
fledg_plot
# Google? y/n [n]
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer
# There seems to be no relationhsip between brightness and fledging success.The
# graph depicts no discernable impact of brightness on fledging success.

bright_plot <- bright_swa %>%
  ggplot(aes(brightness, prop_fledged, color = treatment)) +
  geom_jitter()
bright_plot
# Google? y/n [n]
```
