---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
library(tidyverse)

```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer 
swallows <- read.csv("../quiz/swallows.csv")  
brightness <- read.csv("../quiz/brightness.csv")


# Google? n
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Answer

combined <- merge(swallows, brightness, by = "nestbox") # creates the combined data frame
combined # here I ran this to check if the brightness column was included in the data frame
# you can also look in the environment section if "combined" is a data fram with 42 observations of 5 variables

# Google? y
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer
str(combined)  # see whether each column is a character, integer, etc
unique(combined$treatment)  # make sure there are no typos!
sort(combined$brood) # visually see if there are any numbers that are negative or huge
sort(combined$n_fledged) # same as above

# Google n
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer
combined <- mutate(combined, prop_fledged = n_fledged/brood)
combined # check that it looks good

# Google? n
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 
mean_prop <- combined %>%
  filter(prop_fledged != "NaN") %>%
  summarize(average = mean(prop_fledged))
mean_prop
# The mean proportion is 0.5059829


# Google? n
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer
fledgingtreatment <- ggplot(data = combined, aes(x = treatment, y = prop_fledged)) +
  labs(title = "Differences in Fledging success due to predation", x = "Treatment", y = "Proportion of brood reaching adulthood") +
  geom_boxplot() 
  
fledgingtreatment

# Google? n
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer

fledgingbrightness <- ggplot(combined, aes(x=brightness, y=prop_fledged)) +
  labs(title = "Differences in Fledging success due to predation", x = "Female Brightness", y = "Proportion of brood reaching adulthood") +
  geom_point() +
  geom_smooth(method=lm, se = FALSE, linetype = "dashed", color = "darkblue")

fledgingbrightness    # There is no evidence that brightness influences fledging success


# Google? y
```
