---
title: "Quiz 1: Part 2"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
library(ggplot2)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
# Answer 
brightness <- read.csv("../quiz/brightness.csv")
swallows <- read.csv("../quiz/swallows.csv")

# Google? n
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Answer
birds <- merge(brightness, swallows)
# Google? y
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
# Answer
str(birds)
head(birds)

# Google n
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#Answer
prop_fledged <- mutate(birds, rate = n_fledged/brood*100)


# Google? n
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
# Answer 
mean(prop_fledged$rate, na= TRUE)


# Google? y
```

Create a plot showing fledging success differences between treatments

```{r}
#Answer
num_treatment<- as.numeric(birds$treatment)
prop_fledged |> ggplot() + geom_point(aes(num_treatment, prop_fledged$rate))

# Google? y
```

Is there any evidence that female brightness influenced fledging success? Create a plot to visually investigate this question.

```{r}
#Answer
#No, there is no evidence
birds |> ggplot() + geom_point(aes(brightness, prop_fledged$rate))
# Google? n
```
