---
title: "Quiz 1: Part 2"
author: "Gena Sandoval"
format: html
editor: visual
embed-resources: true
---

## Part 2: Coding in action

### Instructions

Please answer these questions in R using whatever syntax makes sense to you. Feel free to use the R help pages and/or notes from class.

**If you use Google or ChatGPT etc. to help you get to an answer, please note so on your answer**

Submit your answers as an html file. Name it "lastname_quiz1.html". Two submission options:\
1. Email to mcnew\@arizona.edu 2. Create a branch in your forked-and-cloned class repo. Add the html file to the student_contributions/ folder. Submit a pull request to Sabrina (extra credit).

```{r, message = F}
# load packages and data 
library(dslabs)
library(dplyr)
```

#### Load your data:

Load the *swallows.csv* into R, call it swallows. This data frame lists results of an experiment on tree swallows. Each row is a nest. Nests were put into one of two **treatments**: simulated predation (predator), or control (control). Along with treatment you have the following columns:\
**nest_fate** = whether the nest fledged any young or whether they all died\
**brood** = number of nestlings\
**n_fledged** = number of nestlings that fledged.

You also have a separate data frame called *brightness.csv* where each row lists the plumage brightness (a sexual signal) for the female at each nest. Load this one up too.

```{r}
#Load package to read files
library(readr)
#Import dataset assigning it a variable name
#Then open dataset to make sure it imported properly
swallows <- read_csv("../quiz/swallows.csv")
View(swallows)
brightness <- read_csv("../quiz/brightness.csv")
View(brightness)

# Google? N
# RStudio is superhelpful in which you can just click on the dataset and it will import it using the commands above.
```

#### Wrangle your data

You realize that you want to add brightness information onto your main swallows data frame. Join these data frames using nest_id as your key variable. Describe how you would check to make sure this join worked as intended.

```{r}
#Create new dataframe assigning it a descriptive variable name
#Opted not to alter the swallows dataset in case I mess up and need the unaltered version again
#Then open dataframe to make sure it merged properly.Thankfully it did. It added brightness according to nestbox.
updated_swallows <- merge(swallows, brightness, by="nestbox")
View(updated_swallows)

# Google? Y
```

Inspect your combined data frame. Do some sort of quick sanity check on each column to catch any obvious entry errors. Correct as necessary

```{r}
#Chose random nestbox IDs and checked each dataframe manually to make sure that the information was stored correctly

# Google N
```

Create a new column called prop_fledged that contains the proportion of nestlings that fledged from each nest.

```{r}
#I did n_fledged divided by brood to obtain the proportion of nestlings that fledged. There was no nest_fate column as that is data from the swallow_nestlings.csv.

#So it takes the value on the right and puts it into the new column identified on the left
updated_swallows$prop_fledged <- updated_swallows$n_fledged / updated_swallows$brood
View (updated_swallows)

# ChatGPT
```

#### Explore the data

What was the mean proportion of nestlings fledged in each treatment?

```{r}
#This was cool because I learned that piping in R and Linux is different. In Linux I'm used to pipe being '|' but in R, '%>%' is used to string commands together. 

mean_prop <- updated_swallows %>%
  group_by(treatment) %>% 
  summarise(mean_prop_fledged = mean(prop_fledged, na.rm = TRUE))
print(mean_prop)

# ChatGPT
```

Create a plot showing fledging success differences between treatments

```{r}
#Load package
library(ggplot2)

#Make Bar plot
#Had issues with color showing up but it was just because I mislabeled the scale_fill values which I have since fixed.
ggplot(mean_prop, aes(x = treatment, y = mean_prop_fledged, fill = treatment)) +
  geom_bar(stat = "identity") +
  labs(title = "Fledging Success by Treatment",
       x = "Treatment",
       y = "Mean Proportion of Fledged Nestlings") +
  theme_minimal() +
  scale_fill_manual(values = c("Control" = "blue", "Predation" = "red")) +
  geom_text(aes(label = round(mean_prop_fledged, 2)), vjust = -0.5)

# ChatGPT
```

Is there any evidence that female brightness influenced fledgling success? Create a plot to visually investigate this question.

```{r}
# Scatter plot with regression line
ggplot(updated_swallows, aes(x = brightness, y = prop_fledged, color = treatment)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Influence of Female Brightness on Fledging Success",
       x = "Female Brightness",
       y = "Proportion of Nestlings Fledged") +
  theme_minimal() +
  scale_color_manual(values = c("Control" = "blue", "Predation" = "red")) +
  theme(legend.title = element_blank())

#Based on the plot it does not seem that there is a strong connection between fledgling success and female plumage brightness. In the previous figure there was a significant decrease in average fledgling survival under predation compared to the control group. This delineation can still be seen in this last plot but there is not a clear association between brightness and offspring survival.

# ChatGPT
```
