---
title: "njm_rot1"
format: html
editor: visual
---

# Rotation 1: Exploring links between maternal microbiota and reproductive outcomes in wild red squirrels (*Tamiasciurus hudsonicus*)

## By: Natalie J. Morris

## *Modeling and plotting red squirrel reproductive phenotypes and alpha microbial diversity across reproductive status and mast years*

[Githib Repository](https://github.com/nj-morris/Rotation1)

## Load packages, import and organize data

```{r, warning=FALSE, message=FALSE}
# library packages
library(dplyr)
library(lme4)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(gtsummary)
library(kableExtra)


# Read in metadata with alpha diversity values
alphaData <- read.csv("rs_micro_alphaData.csv")

# Reorder factors to biological order
alphaData$season <- factor(alphaData$season, levels=c("early spring", "late spring", "summer", "fall"))
alphaData$rep_status <- factor(alphaData$rep_status, levels=c("non-breeding", "pregnant", "lactating"))
alphaData$mast.x <- factor(alphaData$mast.x, levels=c("non-mast", "mast"))

# Designate as factors
alphaData$squirrel_id.x <- as.factor(alphaData$squirrel_id.x)
alphaData$litter_id <- as.factor(alphaData$litter_id)
alphaData$year.x <- as.factor(alphaData$year.x)
alphaData$mast.x <- as.factor(alphaData$mast.x)

# Filter data by reproductive status
  # alphaDataA for modeling date of birth and litter size at birth
    # Non-Breeding/Pregnancy microbial diversity
alphaDataA <- alphaData %>% 
  filter(rep_status != "lactating")
  # alphaDataB for modeling average growth rate and juvenile survival
      # Pregnancy/Lactation microbial diversity
alphaDataB <- alphaData %>% 
  filter(rep_status != "non-breeding")

# Set ggplot theme
theme_set(theme_bw())
```

## Modeling + Plotting

### Birth Date of Litter

#### Birth Date of Litter: Model

```{r}
# Linear mixed model with random effects, three way interaction
mod_shannon_bd <- lmer(bdate ~ scale(ShannonT)*rep_status*mast.x + mother_grid + season + scale(spr_density) + scale(mother_age) + (1|squirrel_id.x) + (1|year.x), data = alphaDataA) 

# Check residuals
qqnorm(resid(mod_shannon_bd))

summary(mod_shannon_bd, ddf = "Kenward-Roger")

# Formatted lmm output table
tbl_regression(mod_shannon_bd, exponentiate = FALSE, pvalue_fun = label_style_pvalue(digits = 3),
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  italicize_levels()
```

```{r}
# Look at pairwise relationships of model
emm_bdS <- emmeans(mod_shannon_bd, pairwise ~ rep_status * mast.x)
emm_bdS

# Formatted emmeans output table
contrast_df <- as.data.frame(emm_bdS$contrasts)
contrast_df %>%
  kable(caption = "Pairwise Comparisons of `emmeans`") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Birth Date of Litter: Plot

```{r}
# Extract model predictions for birth date
alphaDataA$predicted_bdate <- predict(mod_shannon_bd)  # Includes random effects

# Set colors for alphaDataA
spring_palette <- c("#56B4E9", "#009E73")

# Plot model predictions
ggplot(alphaDataA, aes(x = scale(ShannonT), y = bdate, color = rep_status, fill = rep_status)) +
  geom_point(alpha = 0) +  # remove observed data points
  geom_smooth(method = "lm", aes(y = predicted_bdate), 
              se = TRUE, linewidth = 1, linetype = "solid", alpha = 0.2) +  # Trend line with CI
  facet_wrap(~mast.x, labeller = labeller(mast.x = c("non-mast" = "Non-Mast", "mast" = "Mast"))
  ) +  # Facet by mast.x levels
  scale_color_manual(values = spring_palette, name = "Reproductive Status") +  # Spring colors for lines
  scale_fill_manual(values = spring_palette, name = "Reproductive Status") +   # Matching spring colors for 
  labs(
    x = "Shannon Diversity Index (Scaled (Tukey))",
    y = "Birth Date (Predicted)",
  ) +
  theme_minimal(base_size = 14) +  # Minimal theme with larger text
  theme(
    legend.position = "bottom",  # Move legend 
    legend.title = element_text(size = 12, face = "bold"),  # Bold legend title
    legend.text = element_text(size = 12),  # Adjust legend text size
    strip.text = element_text(size = 12, face = "bold"),  # Bold facet labels
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
    axis.title = element_text(size = 14),  # Adjust axis title size
    axis.text = element_text(size = 12), # Adjust axis text size 
    panel.grid.minor = element_blank(), panel.grid.major = element_blank(),
    axis.line = element_line(color = "black", linewidth  = 0.5), 
    panel.border = element_rect(color = "black", linewidth  = 0.5, fill = NA)  # Add border between facets
    
  )
```

#### Birth Date of Litter: Interpretation

Earlier dates of birth were predicted by greater microbial diversity, which did not differ across reproductive status or mast years.

### Litter Size at Birth

#### Litter Size at Birth: Model

```{r}
# Linear mixed model with random effects, three way interaction
mod_shannon_ls <- lmer(litter_size ~ scale(ShannonT)*rep_status*mast.x + mother_grid + season
                       + scale(spr_density) + scale(mother_age) + (1|squirrel_id.x) + (1|year.x), data = alphaDataA) 

# Check residuals 
qqnorm(resid(mod_shannon_ls))

summary(mod_shannon_ls, ddf = "Kenward-Roger")

# Formatted lmm output table
tbl_regression(mod_shannon_ls, exponentiate = FALSE, pvalue_fun = label_style_pvalue(digits = 3),
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  italicize_levels()
```

```{r}
# Look at pairwise relationships of model
emm_lsS <- emmeans(mod_shannon_ls, pairwise ~ rep_status * mast.x)
emm_lsS

# Formatted emmeans output table
contrast_df <- as.data.frame(emm_bdS$contrasts)
contrast_df %>%
  kable(caption = "Pairwise Comparisons of `emmeans`") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Litter Size at Birth: Plot

```{r}
# Extract model predictions for litter size
alphaDataA$predicted_ls <- predict(mod_shannon_ls)  # Includes random effects

# Set colors for alphaDataA
spring_palette <- c("#56B4E9", "#009E73")

# Plot model predictions
ggplot(alphaDataA, aes(x = scale(ShannonT), y = litter_size, color = rep_status, fill = rep_status)) +
  geom_point(alpha = 0) +  # Observed data points
  geom_smooth(method = "lm", aes(y = predicted_ls), 
              se = TRUE, linewidth = 1, linetype = "solid", alpha = 0.2) +  # Trend line with CI
  facet_wrap(~mast.x, labeller = labeller(mast.x = c("non-mast" = "Non-Mast", "mast" = "Mast"))) +  # Facet by mast.x levels
  scale_color_manual(values = spring_palette, name = "Reproductive Status") +  # Spring colors for lines
  scale_fill_manual(values = spring_palette, name = "Reproductive Status") +   # Matching spring colors for CI
  labs(
    x = "Shannon Diversity Index (Scaled (Tukey))",
    y = "Litter Size (Predicted)",
  ) +
  theme_minimal(base_size = 14) +  # Minimal theme with larger text
  theme(
    legend.position = "bottom",  # Move legend 
    legend.title = element_text(size = 12, face = "bold"),  # Bold legend title
    legend.text = element_text(size = 12),  # Adjust legend text size
    strip.text = element_text(size = 12, face = "bold"),  # Bold facet labels
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
    axis.title = element_text(size = 14),  # Adjust axis title size
    axis.text = element_text(size = 12), # Adjust axis text size 
    panel.grid.minor = element_blank(), panel.grid.major = element_blank(),
    axis.line = element_line(color = "black", linewidth  = 0.5), 
    panel.border = element_rect(color = "black", linewidth  = 0.5, fill = NA)  # Add border between facets
)
```

#### Litter Size at Birth: Interpretation

There was not a strong trend of alpha diversity predicting litter size, however this relationship differed across mast and non-mast years, but not reproductive status.

### Growth Rate of Litter (Avg)

#### Growth Rate of Litter: Model

```{r}
# Linear mixed model with random effects, three way interaction
mod_shannon_gr <- lmer(avg_growth_litter ~ scale(ShannonT)*rep_status*mast.x + mother_grid + season
                       + scale(spr_density) + scale(mother_age) + (1|squirrel_id.x) + (1|year.x), data = alphaDataB) 

# Check residuals
qqnorm(resid(mod_shannon_gr))

summary(mod_shannon_gr, ddf = "Kenward-Roger")

# Formatted lmm output table
tbl_regression(mod_shannon_gr, exponentiate = FALSE, pvalue_fun = label_style_pvalue(digits = 3),
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  italicize_levels()
```

```{r}
# Look at pairwise relationships of model
emm_grS <- emmeans(mod_shannon_gr, pairwise ~ rep_status * mast.x)
emm_grS

# Formatted emmeans output table
contrast_df <- as.data.frame(emm_grS$contrasts)
contrast_df %>%
  kable(caption = "Pairwise Comparisons of `emmeans`: Growth Rate") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Growth Rate of Litter: Plot

```{r}
# Extract model predictions for growth rate
alphaDataB$predicted_gr <- NA  # Initialize the column with NA
alphaDataB$predicted_gr[complete.cases(alphaDataB[c("avg_growth_litter", "ShannonT", "rep_status", "mast.x", "mother_grid", "season", "spr_density", "mother_age")])] <- predict(mod_shannon_gr)

# Specify colors for alphaDataB
spring_palette <- c("#009E73", "#CC79A7" )

# Plot model predictions
ggplot(alphaDataB, aes(x = scale(ShannonT), y = avg_growth_litter, color = rep_status, fill = rep_status)) +
  geom_point(alpha = 0) +  # Observed data points
  geom_smooth(method = "lm", aes(y = predicted_gr), 
              se = TRUE, linewidth = 1, linetype = "solid", alpha = 0.2) +  # Trend line with CI
  facet_wrap(~mast.x, labeller = labeller(mast.x = c("non-mast" = "Non-Mast", "mast" = "Mast"))) +  # Facet by mast.x levels
  scale_color_manual(values = spring_palette, name = "Reproductive Status") +  # Spring colors for lines
  scale_fill_manual(values = spring_palette, name = "Reproductive Status") +   # Matching spring colors for CI
  labs(
    x = "Shannon Diversity Index (Scaled (Tukey))",
    y = "Growth Rate (Predicted)",
  ) +
  theme_minimal(base_size = 14) +  # Minimal theme with larger text
  theme(
    legend.position = "bottom",  # Move legend 
    legend.title = element_text(size = 12, face = "bold"),  # Bold legend title
    legend.text = element_text(size = 12),  # Adjust legend text size
    strip.text = element_text(size = 12, face = "bold"),  # Bold facet labels
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
    axis.title = element_text(size = 14),  # Adjust axis title size
    axis.text = element_text(size = 12), # Adjust axis text size 
    panel.grid.minor = element_blank(), panel.grid.major = element_blank(),
    axis.line = element_line(color = "black", linewidth  = 0.5), 
    panel.border = element_rect(color = "black", linewidth  = 0.5, fill = NA)  # Add border between facets
  )

```

#### Growth Rate: Interpretation

The trend shows faster growth rate (averaged) of litters with lower microbial diversity in non-mast years at both pregnant and lactating sampling time points, but this relationship diverged in mast years, where faster growth rate was predicted by greater microbial diversity during pregnancy.

### Juvenile Survival (survived first winter)

#### Juvenile Survival: Model

```{r}
# Linear mixed model with random effects, three way interaction
mod_shannon_jsLMM <-lmer(juv_survived ~ scale(ShannonT)*rep_status*mast.x + scale(mother_age) + mother_grid +
                            scale(spr_density) + season + offset(log(litter_size)) + (1|squirrel_id.x) + (1|year.x), data = alphaDataB)

# Check residuals
qqnorm(resid(mod_shannon_jsLMM))

summary(mod_shannon_jsLMM)

# Formatted lmm output table
tbl_regression(mod_shannon_jsLMM, exponentiate = FALSE, pvalue_fun = label_style_pvalue(digits = 3),
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  italicize_levels()
```

```{r}
# Look at pairwise relationships of model
emm_jsS <- emmeans(mod_shannon_jsLMM, pairwise ~ rep_status * mast.x)
emm_jsS

emmeans(mod_shannon_jsLMM, ~ scale(ShannonT) | rep_status * mast.x)


# Formatted emmeans output table
contrast_df <- as.data.frame(emm_jsS$contrasts)
contrast_df %>%
  kable(caption = "Pairwise Comparisons of `emmeans`: Juvenile Survival") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

#### Juvenile Survival: Plot

```{r}
# Extract model predictions for juvenile survival
alphaDataB$predicted_js <- NA  # Initialize the column with NA
alphaDataB$predicted_js[complete.cases(alphaDataB[c("juv_survived", "ShannonT", "rep_status", "mast.x", 
                                                  "mother_grid", "season", "spr_density", "mother_age")])] <- 
  predict(mod_shannon_jsLMM)

# Plot model predictions
ggplot(alphaDataB, aes(x = scale(ShannonT), y = juv_survived, color = rep_status, fill = rep_status)) +
  geom_point(alpha = 0) +  # Observed data points
  geom_smooth(method = "lm", aes(y = predicted_js), 
              se = TRUE, linewidth = 1, linetype = "solid", alpha = 0.2) +  # Trend line with CI
  facet_wrap(~mast.x, labeller = labeller(mast.x = c("non-mast" = "Non-Mast", "mast" = "Mast"))) +  # Facet by mast.x levels
  scale_color_manual(values = spring_palette, name = "Reproductive Status") +  # Spring colors for lines
  scale_fill_manual(values = spring_palette, name = "Reproductive Status") +   # Matching spring colors for CI
  labs(
    x = "Scaled(Shannon Diversity Index (Tukey))",
    y = "Juvenile Survival (Predicted)",
  ) +
  theme_minimal(base_size = 14) +  # Minimal theme with larger text
  theme(
    legend.position = "bottom",  # Move legend 
    legend.title = element_text(size = 12, face = "bold"),  # Bold legend title
    legend.text = element_text(size = 12),  # Adjust legend text size
    strip.text = element_text(size = 12, face = "bold"),  # Bold facet labels
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
    axis.title = element_text(size = 14),  # Adjust axis title size
    axis.text = element_text(size = 12), # Adjust axis text size 
    panel.grid.minor = element_blank(), panel.grid.major = element_blank(),
    axis.line = element_line(color = "black", linewidth  = 0.5), 
    panel.border = element_rect(color = "black", linewidth  = 0.5, fill = NA)  # Add border between facets
  )
```

#### Juvenile Survival: Interpretation

There was a trend of greater juvenile survival with lower microbial alpha diversity during pregnancy and greater microbial alpha diversity during lactation. This relationship did not significantly differ across non-mast and mast years, but was significantly different across reproductive status only in mast years.

### Alpha diversity over sampling timeline: Plot

```{r}
# Plot raw alpha diversity across days since parturition of each individual reproductive attempt #
alphaData %>%
  ggplot(aes(x = days_since_part, y = Shannon)) +
  geom_line(aes(group = litter_id, color = litter_id), alpha = 0.75) +                  # Individual trajectories with transparency
  geom_point(aes(color = litter_id), alpha = 0.5) +                                    # Individual data points with transparency
  geom_smooth(aes(group = 1),                                                          # Overall trend line across all individuals
              method = "loess", 
              color = "black", 
              se = TRUE,                    # Add confidence interval around trend line
              linetype = "dashed") +
  scale_color_viridis_d() +  
  # Add shaded ranges
  annotate("rect", xmin = -89, xmax = -35, ymin = -Inf, ymax = Inf, fill = "#56B4E9", alpha = 0.1) + # Non-Breeding phase
  annotate("rect", xmin = -35, xmax = 0, ymin = -Inf, ymax = Inf, fill = "#009E73", alpha = 0.1) + # Parturition phase
  annotate("rect", xmin = 0, xmax = 70, ymin = -Inf, ymax = Inf, fill = "#CC79A7", alpha = 0.1) + # Lactating phase
  annotate("rect", xmin = 71, xmax = Inf, ymin = -Inf, ymax = Inf, fill = "#56B4E9", alpha = 0.1) + # Lactating phase
  # Add labels for ranges
  annotate("text", x = -60, y = max(alphaData$Shannon) * 1.1, label = "Non-Breeding", color = "#56B4E9") +
  annotate("text", x = -20, y = max(alphaData$Shannon) * 1.1, label = "Pregnant", color = "#009E73") +
  annotate("text", x = 35, y = max(alphaData$Shannon) * 1.1, label = "Lactating", color = "#CC79A7") +
  annotate("text", x = 85, y = max(alphaData$Shannon) * 1.1, label = "Non-Breeding", color = "#56B4E9") +
  labs(x = "Days Since Parturition", 
       y = "Shannon Gut Microbial Diversity", 
       title = "Shannon Diversity Across Days Since Birth") +
  theme_minimal() +
  theme(legend.position = "none",
        legend.title = element_text(size = 12, face = "bold"),  # Bold legend title
        legend.text = element_text(size = 12),                 # Adjust legend text size
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
        axis.title = element_text(size = 14), 
        axis.text = element_text(size = 12),  
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(),
        axis.line = element_line(color = "black", linewidth  = 0.5))
```

#### Alpha diversity over sampling timeline: Interpretation

Across individuals and days since parturition (birth), alpha diversity varied. There was no clear strategy–some individuals' microbial diversity decreased and others increased, while some remained largely stable and others changed for portions of the sampling timeline.
